\bf Simulation-guided Data Assimilation for Kinetics Modeling of Potassium Channel Isoforms in Mouse Cardiomyocytes

### 2022-06-05 writing/draft.tex
Method section completed & New intro

\noindent%
{\it Keywords:} Potassium channels, kinetics modeling, data assimilation, model calibration, reduced glycosylation

@@

\section{Introduction} \label{s:intro}
\begin{itemize}
    \item \textbf{Motivation:} Aberrant activities of voltage-gated potassium (K\textsubscript{v}) channels can cause fatal heart disease such as long QT syndrome. In our recent investigation on a disease-related perturbation, reduced glycosylation in cardiomyocytes, significant reduction in K\textsuperscript{+} currents was observed. Among major cardiac voltage-gated ion channels (VGIC), K\textsubscript{v} channels have a distinctive feature: various isoforms generate their own unique currents contributing to different phases of repolarization of cardiac action potential (AP). On the other hand, the dominant isoform is responsible for Na\textsuperscript{+} and Ca\textsuperscript{2+} currents. Because only the sum of these K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be recorded in experiments, data analytical methods are necessary to estimate the properties of individual K\textsuperscript{+} currents and their effects on the behaviors of the cardiomyocyte.
    \item \textbf{Gaps \& Needs:} However, traditional approaches estimate only one property, the shape of individual K\textsuperscript{+} currents, from a single I\textsubscript{Ksum} recording rather than analyzing I\textsubscript{Ksum} recordings from the same cell with different conditions. These approaches assume the functional form of the shape of K\textsuperscript{+} currents and search the shape parameters in the current functions that the sum of estimated currents has the best fit with a single I\textsubscript{Ksum} recording. It does not provide information about underlying kinetics of the currents at the cellular level.
    \item \textbf{Objective:} In this paper, we present a novel framework of data assimilation method that delineates kinetics of K\textsubscript{v} isoforms from I\textsubscript{Ksum} recordings based on computer models of K\textsuperscript{+} currents that simulate underlying biophysics of the currents. As opposed to the traditional data-driven curve fitting method, the proposed biophysics-driven method estimates the cellular-level kinetics of K\textsubscript{v} isoforms.
    \item \textbf{Approach:} First, computer models of K\textsuperscript{+} currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. Second, we develop fractional factorial designs to identify the parameters that have significant impacts on the current shapes. Third, we design model calibration as nonlinear optimization that couple the in-silico model and in-vivo I\textsubscript{Ksum} recordings by minimizing discrepancies between them. Further, we develop a graphical user interface (GUI) application to make the proposed framework accessible for researchers and patricians without engineering backgrounds.
    \item \textbf{Experiment \& Performance} We apply the proposed framework to our data, and the experimental results show that the data assimilation method can estimate differences in kinetic rate between the control and experiment groups.
    \item \textbf{Significance/Contributions} 1) We model the underlying kinetics of observed currents based on biophysical principles rather than curve-fitting methods. 2) The calibration process includes multiple protocols rather than a single protocol, resulting in comprehensive cellular-level modeling. 3) Our method provides differences in channel kinetics given the disease-related perturbation, such as reduced glycosylation. 
\end{itemize}

\begin{figure}
    \centering
    \includegraphics{figs/overall_diagram2.pdf}
    \caption{Caption}
    \label{fig:framework_diagram}
\end{figure}

@@

\subsection{\emph{Sensitivity Analysis of Calibration Parameters}} \label{s:methods.3}
The principle of parsimony is critical in model calibration not only for enhancing fitting accuracy and preventing overfitting, but also for improving the interpretability of $\theta$. We performed a sensitivity analysis to identify a subset of the parameters that have significant impacts on the model output, i.e., current traces. We defined six \textit{anchor points} that capture representative characteristics of K\textsuperscript{+} current traces in voltage-clamp experiments as illustrated in \textbf{Figure~\ref{fig:current_feature}}. Each point represents: a) the current magnitude of 10 ms after applying a voltage step, which measures the activation rate; b) 25\% of the total recording time has elapsed, c) 50\%, and d) 75\%, which collectively estimate the inactivation rate; e) the peak magnitude; and f) the time when current has decayed ($1-e^{-1}$)\% (almost 63\%) from the peak. Feature f will be equal to the total recording time if current does not decline enough as in \textbf{Figure~\ref{fig:current_feature}C} and \textbf{~\ref{fig:current_feature}D}.

We developed factorial designs in which parameters varied at two levels, contrasting their effect. The anchor points were evaluated at each design to calculate factorial effects. Then, half-normal plots were drawn based on these contrasts, providing visual inspections of critical parameters that exert the most impact on each shape feature. All the half-normal plots can be found in Supplement A. \textbf{Table~\ref{tab:calib_param}} summarizes the selected parameters that will be calibrated to couple the \textit{in-silico} models with \textit{in-vitro} experimental data. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/anchor_ponts.pdf}
    \caption{Examples of the six feature of K\textsuperscript{+} currents that quantify characteristics of the current shape of (A) I\textsubscript{Kto}, (B) I\textsubscript{Kslow1}, (C) I\textsubscript{Kslow2}, and (D) I\textsubscript{Kss}. All currents are simulated for illustration, and the feature labels refer to a: the current magnitude 10 ms after voltage is applied, b: 25\% of the total recording time has elapsed, c: 50\%, d: 75\%, e: the peak magnitude, and f: the time when current has decayed ($1-e^{-1}$)\% (almost 63\%) from the peak.}
    \label{fig:current_feature}
\end{figure}
\begin{table}
    \centering
    \caption{Calibration parameters that will be optimized to minimize the discrepancy between model predictions and experimental data.}
    \begin{tabular}{l|l}
        \hline
        Model & Calibration parameters \\
        \hline
        K\textsubscript{v}4.2 (I\textsubscript{Kto}) & $p_{1},\quad p_{2},\quad p_{4},\quad p_{5},\quad p_{7},\quad p_{11},\quad G_{\mathrm{Kto}}$ \\
        K\textsubscript{v}1.5 (I\textsubscript{Kslow1}) & $p_{1},\quad p_{2},\quad p_{4},\quad p_{5},\quad p_{9},\quad p_{10},\quad G_{\mathrm{Kslow1}}$ \\
        K\textsubscript{v}2.1 (I\textsubscript{Kslow1}) & $p_{2},\quad G_{\mathrm{Kslow2}}$ \\
        K\textsubscript{2P} (I\textsubscript{Kss}) & $p_{1},\quad p_{2},\quad p_{3},\quad G_{\mathrm{Kss}}$ \\
        \hline
    \end{tabular}
    \label{tab:calib_param}
\end{table}

We categorized the calibration parameters into four classes according to their functional roles in kinetic functions, which are highlighted in different colors in \textbf{Figure~\ref{fig:calib_param}}. 
red: voltage threshold, green: voltage slope, blue: kinetic-function factor, and time-constant shift. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/calib_param.pdf}
    \caption{Calibration parameters with highlights according to the class}
    \label{fig:calib_param}
\end{figure}

\subsection{\emph{Glycosylation and Dilated Cardiomyopathy (DCM)}} \label{s:methods.4}
Case study with our own in-vitro experimental data.

\section{Results} \label{s:results}
\subsection{Experimental Design: Reduced Glycosylation}
case study with the in-vitro data of investigating the casuality between reduced glycosylation and dilated cardiomyopathy.

\subsection{Experimental Results}
case study results with the proposed data assimilation method

Perhaps numerical analysis of performance, comparison, and demonstration of applicability and impact using real data.

>>>>

\noindent%
{\it Keywords:} Data assimilation, potassium channels, kinetics modeling, model calibration, glycosylation, cardiomyopathy,

@@

\section{Introduction} \label{s:intro}
Cardiac voltage-gated potassium channels (K\textsubscript{v}) are responsible for repolarizing the action potential (AP) in cardiomyocytes. They are indispensable in the electrical conduction system of the heart. Even modest changes in K\textsubscript{v} activities can significantly affect the AP duration and the QT interval, which lead to fatal heart diseases \citep{ravens2008role}. Glycosylation is a co/posttranslational modification that is critical for protein functions including activities of voltage-gated ion channels (VGICs) \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies in diverse directions, ranging from genome-wise to proteomic/glycomic searches, has shown clear links between altered protein glycosylation and heart diseases \citep{yung2004gene,yang2015glycoproteins,miura2016glycomics}. However, the underlying mechanism is still largely unknown how changes in glycosylation contribute to heart disease onset and progression. Data from our \textit{in-vitro} experiments show that glycosylation regulations modulate VGIC activities and contribute to both electrical and contractile dysfunction \citep{ednie2013sialicNav1,ednie2015sialicKv,ednie2019reduced}. We also conducted in-silico experiments to verify the data and investigate how changes in the lower level (VGICs) affect the functions in the higher level (cardiomyocytes), which is one of the unique powers of systematic computer models \citep{du2013silico,du2015statistical,du2017silico}. As illustrated in \textbf{Figure~\ref{fig:framework_diagram}}, 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/overall_diagram2.pdf}
    \caption{Caption}
    \label{fig:framework_diagram}
\end{figure}

However, unique feature of Kv raise challenges.
\begin{itemize}
    \item \textbf{Gaps \& Needs:} Among major cardiac voltage-gated ion channels (VGIC), K\textsubscript{v} channels have a distinctive feature: various isoforms generate their own unique currents contributing to different phases of repolarization of cardiac action potential (AP). On the other hand, the dominant isoform is responsible for Na\textsuperscript{+} and Ca\textsuperscript{2+} currents. Because only the sum of these K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be recorded in experiments, data analytical methods are necessary to estimate the properties of individual K\textsuperscript{+} currents and their effects on the behaviors of the cardiomyocyte. However, traditional approaches estimate only one property, the shape of individual K\textsuperscript{+} currents, from a single I\textsubscript{Ksum} recording rather than analyzing I\textsubscript{Ksum} recordings from the same cell with different conditions. These approaches assume the functional form of the shape of K\textsuperscript{+} currents and search the shape parameters in the current functions that the sum of estimated currents has the best fit with a single I\textsubscript{Ksum} recording. It does not provide information about underlying kinetics of the currents at the cellular level.
    \item \textbf{Objective:} In this paper, we present a novel framework of data assimilation method that delineates kinetics of K\textsubscript{v} isoforms from I\textsubscript{Ksum} recordings based on computer models of K\textsuperscript{+} currents that simulate underlying biophysics of the currents. As opposed to the traditional data-driven curve fitting method, the proposed biophysics-driven method estimates the cellular-level kinetics of K\textsubscript{v} isoforms.
    \item \textbf{Approach: (need a flowchart} First, computer models of K\textsuperscript{+} currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. Second, we develop fractional factorial designs to identify the parameters that have significant impacts on the current shapes. Third, we design model calibration as nonlinear optimization that couple the in-silico model and in-vivo I\textsubscript{Ksum} recordings by minimizing discrepancies between them. Further, we develop a graphical user interface (GUI) application to make the proposed framework accessible for researchers and patricians without engineering backgrounds.
    \item \textbf{Experiment \& Performance} We apply the proposed framework to our data, and the experimental results show that the data assimilation method can estimate differences in kinetic rate between the control and experiment groups.
    \item \textbf{Significance/Contributions} 1) We model the underlying kinetics of observed currents based on biophysical principles rather than curve-fitting methods. 2) The calibration process includes multiple protocols rather than a single protocol, resulting in comprehensive cellular-level modeling. 3) Our method provides differences in channel kinetics given the disease-related perturbation, such as reduced glycosylation. 
\end{itemize}

@@

\subsection{\emph{Sensitivity Analysis and Model Regularization}} \label{s:methods.3}
The principle of parsimony is critical in model calibration not only for enhancing fitting accuracy and preventing overfitting, but also for improving the interpretability of $\theta$. We performed a sensitivity analysis to identify a subset of the parameters that have significant impacts on the model output and only calibrate these sensitive parameters. As illustrated in \textbf{Figure~\ref{fig:anchor_points}}, six \textit{anchor points} were defined that capture representative characteristics of K\textsuperscript{+} current traces in voltage-clamp experiments. Each point represents: a) the current magnitude of 10 ms after applying a voltage step, which measures the activation rate; b) 25\% of the total recording time has elapsed, c) 50\%, and d) 75\%, which collectively estimate the inactivation rate; e) the peak magnitude; and f) the time when current has decayed ($1-e^{-1}$)\% (almost 63\%) from the peak. Feature f will be equal to the total recording time if current does not decline enough as in \textbf{Figure~\ref{fig:anchor_points}C} and \textbf{~\ref{fig:anchor_points}D}. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/anchor_ponts.pdf}
    \caption{Illustration of the six anchor points of voltage-clamp K\textsuperscript{+} currents that quantify characteristics of the current shape of (A) I\textsubscript{Kto}, (B) I\textsubscript{Kslow1}, (C) I\textsubscript{Kslow2}, and (D) I\textsubscript{Kss}. All currents are simulated for illustration, and the labels refer to a) the current magnitude 10 ms after voltage is applied, b) 25\% of the total recording time has elapsed, c) 50\%, d) 75\%, e) the peak magnitude, and f) the time when current has decayed ($1-e^{-1}$)\% (almost 63\%) from the peak.}
    \label{fig:anchor_points}
\end{figure}

We developed factorial designs in which parameters varied at two levels, contrasting their effect. The anchor points were evaluated at each design to calculate factorial effects. Then, half-normal plots were drawn based on these contrasts, providing visual inspections of critical parameters that exert the most impact on each shape feature. All the half-normal plots can be found in Supplement A. \textbf{Figure~\ref{fig:calib_param}} shows the selected parameters, highlighted in different colors according to their functional roles in channel kinetics. We categorized the calibration parameters into four classes: The red represents the voltage-threshold parameters and the green voltage slopes, controlling the voltage dependence, the blue scale factors of kinetic functions, and purple time-constant shifters. Note that the voltage-dependence parameters in red and green appear multiple times across different equations. This parsimonious model design is intended to maximize the structural regularization to minimize overfitting.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/calib_param.pdf}
    \caption{Selected parameters from the sensitivity analysis, highlighted in different colors according to their functional roles in channel kinetics.}
    \label{fig:calib_param}
\end{figure}

\subsection{\emph{Glycosylation and Dilated Cardiomyopathy (DCM)}} \label{s:methods.4}
We applied the proposed framework to our studies on the pathophysiology of electrical signaling by altered glycosylation. Protein glycosylation is one of the most abundant and diverse forms of co/posttranslational modifications that impact essential protein functions, such as modulation of receptor or ion channel activities \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies has shown the association between altered glycosylation and heart diseases, such as dilated cardiomyopathy (DCM) and hypertrophic cardiomyopathy \citep{ohtsubo2006glycosylation,ednie2019reduced2}. It is reported that up to 20\% of patients with congenital disorders of glycosylation (CDG), who commonly have modest reductions in protein glycosylation, present with cardiac deficits and idiopathic DCM \citep{marques2017cardiac}. We have investigated how regulated glycosylation contributes to heart failure in the context of electrophysiology. Electrical signaling is orchestrated activities of a variety of ion channels and transporters. Voltage-gated ion channels (VGICs) are heavily glycosylated, with ~30\% of the channel mass consisting of N-/O-linked glycans \citep{ednie2012modulation}.

Glycosylation is a multi-step process and usually ends with sialic acid added. We reported that a saturating, electrostatic effect of negatively charged sialic attached to the terminal of N-/O-glycan branches significantly altered electrical signaling in Na\textsubscript{v} \citep{ednie2013sialicNav1,ednie2015sialicNav2} as well as K\textsubscript{v} \citep{ednie2015sialicKv}. More recently, we showed that preventing hybrid/complex N-glycosylation in mouse cardiomyocytes was sufficient to cause DCM using genetic ablation of the \textit{MGAT1} gene (MGAT1KO model), which encodes a critical glycosyltransferase, GlcNAcT1 \citep{ednie2019reduced2}. MGAT1KO mice developed DCM, deteriorated into heart failure, and 100\% died early, presumably from ventricular arrhythmias leading to sudden cardiac death. To further investigate the role of altered glycosylation and pathogenesis and disease progression of the heart, we conducted whole-cell patch-clamp experiments that showed reductions in N-glycosylation significantly impact electrical signaling in mouse cardiomyocytes \citep{ednie2019reduced}. However, it was difficult to determine channel kinetics rigorously from the \textit{in-vitro} experiments alone, especially for K\textsuperscript{+} current recordings, due to vast diversity of K\textsubscript{v} isoforms and small current magnitude of the MGAT1KO model. Therefore, we validate the suggested framework by modeling K\textsubscript{v} isoforms with our \textit{in-vitro} experimental data. Therefore, we leverage the suggested framework to delineate the channel kinetics of K\textsubscript{v} to provide new descriptive/predictive insights into our \textit{in-vitro} experiments.

\section{Results} \label{s:results}

\subsection{Calibration Results}

\subsection{Distributions Estimations}

\subsection{Clustering}
Perhaps numerical analysis of performance, comparison, and demonstration of applicability and impact using real data.

\section{Discuss}
Discuss of the results goes here.
